Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-03 article.tex
Abschnitt 3

@@

A system of ordinary differential equation $\dot x = f(x, p, u, t)$ can be represented by a \code{character} vector of equations involving symbols for the states $x$ (coinciding with names of the vector), symbols for the parameters $p$, symbols for the forcings $u$ and the keyword \code{time} for $t$. The class provided by \pkg{dMod} for such objects is the \code{eqnvec} class which checks whether the character can be parsed and thus, can be interpreted as an equation.

@@

@@

@@

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied, yielding $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is translated into a \proglang{C} code and compiled by default.

@@

In principle, a parameter transformation $p = \Phi(\theta)$ is a (differentiable) function connecting \textbf{inner} parameters $p$ with \textbf{outer} parameters $\theta$. The rationale behind the distinction of inner and outer parameters is that the vector $p$ usually desribes those parameters defined in the model equations. The outer parameters $\theta$ refer to a convenient parameterization by which the model parameters are computed.

An \proglang{R} function of class \code{parfn} is produced by the \code{P()} command. Transformations can either be formulated explicitly or implicitly. In the explicit case, the function $p = \Phi(\theta)$ corresponds to an \code{eqnvec} whose names are the names of the inner parameters and entries are equations with symbols for the outer parameters. An implicit transformation has the form $f(p = \Phi(\theta), \theta) = 0$. In this case, $f$ is expressed by an \code{eqnvec} with equations containing symbols for $p$ and $\theta$ and the names of the \code{eqnvec} are the symbols for $p$.

@@

The aim of \pkg{dMod} is to allow for "simultaneous" prediction under several conditions and compare these predictions to the corresponding experimental data sets to estimate model parameters. The full information contained in several perturbation experiments is only used if the multi-conditional model prediction is set up in a way that all model parameters not being changed from condition to condition correspond to exactly one parameter. On the other hand, parameters which are systematically perturbed correspond to different parameters.

Mathematically speaking, we want to construct a parameter transformation
\begin{equation}
	\begin{aligned}
		\Phi:\quad \mathbb R^p & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^q \\
		\theta & \longmapsto  \left(\Phi_1(\theta), \dots, \Phi_n(\theta)\right)
	\end{aligned}
	\label{eq:partrafo}
\end{equation}
where $i = 1, \dots, n$ corresponds to the different conditions. If all parameters are shared throughout all conditions, then $p = q$. If, however, all parameters are distinct, then $p = n\cdot q$. Perturbation experiments correspond to a situation where $p\gtrsim q$.

The \pkg{dMod} package allows to define transformations $\Phi$, see eq.~\eqref{eq:partrafo}, by the \code{"+"} operator: 
\begin{equation*}
\verb!p <- P(eqnvec1, condition = "one") + P(eqnvec2, condition = "two")!. 
\end{equation*}
All symbols from \code{eqnvec1} and \code{eqnvec2} are collected and their union constitutes the symbols of $\theta$. The evaluation \code{p(theta)} returns a list (of length two in the example) with inner parameters. The \code{"+"} operator can be applied consecutively to add condition by condition.

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)) 
	\end{aligned}
	\label{}
\end{align*}
and the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined and combined by the \code{"+"} operator analogously to \code{p}.

All commands, \code{P()}, \code{Xs()}, \code{Y()}, etc.~can be executed with \code{condition = NULL}. In that case, the corresponding returned function is generic and, if called for different conditions, the same identical function is evaluated with the condition-specific input.

@@

The aim of \pkg{dMod} is parameter estimation. The objective function to be minimized for this purpose is the weighted least squares function which is produced by the command \code{normL2(data, prdfn)} where \code{data} is a \code{datalist} object and \code{prdfn} is a prediction function. A \code{datalist} is a list of \code{data.frame}s providing time points, observables, measured values and uncertainties for each condition. The \code{as.datalist()} command allows to construct a datalist from a \code{data.frame} as it is typically used for mixed effects models or plotting with \pkg{ggplot2}.

>>>>

The guideline of \pkg{dMod} is to provide a class structure for models, predictions, observations and parameter transformations that allows a flexible combination of many experimental conditions in one objective function to fit models to data. This flexibility is achieved by two concepts: (1) concatenation of functions by a \code{"*"} operator and (2) stacking of functions representing different conditions by a \code{"+"} operator. The handling and propagation of derivatives as part of the classes happens in the background. Methods for the generic \code{print()}, \code{plot()} and \code{summary()} functions are implemented for most outputs.

@@

A system of ordinary differential equation $\dot x = f(x, p, u, t)$ can be represented by a \code{character} vector of symbolic expressions, the right-hand side of the ODE, involving symbols for the states $x$, the parameters $p$, the forcings $u$ and the keyword \code{time} for $t$. The class provided by \pkg{dMod} for such objects is the \code{eqnvec} class which checks whether the character can be parsed and thus, can be interpreted as an equation.

@@

\subsection{Prediction functions}

@@

Besides parameters, the prediction might also depend on forcings $u(t)$ and sudden events, e.g.~setting states to 0 at a predefined point in time. Since forcings and events are by default not variable, they are defined for a prediction function the moment it is generated, \code{x <- Xs(odemodel, forcings = myforcings, events = myevents)}. The definitions of both, forcings and events, are compatible with the way they are defined in \pkg{deSolve}.

@@

% Concatenation in Einleitungssatz erwähnen.

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y = g\circ x$, i.e.~$y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied, yielding $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is usually translated into a \proglang{C} code and compiled.

@@

In principle, a parameter transformation $p = \Phi(\theta)$ is a (differentiable) function connecting \textbf{inner} parameters $p$ with \textbf{outer} parameters $\theta$. The rationale behind the distinction of inner and outer parameters is that the vector $p$ usually desribes those parameters defined in the model equations. The outer parameters $\theta$ refer to a convenient parameterization by which the model parameters are computed. Examples are a log-transform of the inner parameters, $\theta = \log(p) \Leftrightarrow p = \Phi(\theta) = e^{\theta}$, or parameter constraints like $(p_1, p_2) = (\theta_1, \theta_1 + \theta_2)$.

An \proglang{R} function of class \code{parfn} is produced by the \code{P()} command. Transformations can either be formulated explicitly or implicitly. In the explicit case, the function $p = \Phi(\theta)$ corresponds to an \code{eqnvec} whose names are the names of the inner parameters and entries are equations with symbols for the outer parameters. An implicit transformation has the form $f\big(p = \Phi(\theta), \theta\big) = 0$. In this case, $f$ is expressed by an \code{eqnvec} with equations containing symbols for $p$ and $\theta$ and the names of the \code{eqnvec} are the symbols for $p$.

@@

The aim of \pkg{dMod} is to allow for "simultaneous" predictions under several conditions and compare these predictions to the corresponding experimental data sets to estimate model parameters.
Different experimental conditions can be expressed by that fact that some parameters are different between conditions whereas others are common to all conditions. This situation occurs e.g.~if perturbation experiments are performed, affecting only few parts of a system.
%The full information contained in several perturbation experiments is only used if the multi-conditional model prediction is set up in a way that all model parameters not being changed from condition to condition correspond to exactly one parameter. On the other hand, parameters which are systematically perturbed correspond to different parameters.
Mathematically speaking, we want to construct a parameter transformation
\begin{equation}
	\begin{aligned}
		\Phi:\quad \mathbb R^p & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^q \\
		\theta & \longmapsto  \left(\Phi_1(\theta), \dots, \Phi_n(\theta)\right)
	\end{aligned}
	\label{eq:partrafo}
\end{equation}
where $i = 1, \dots, n$ corresponds to the different conditions. If all parameters are shared throughout all conditions, then $p = q$. If, however, all parameters are distinct, then $p = n\cdot q$. Perturbation experiments correspond to a situation where $p\gtrsim q$.

The \pkg{dMod} package allows to define transformations $\Phi$, see eq.~\eqref{eq:partrafo}, by the \code{"+"} operator: 
\begin{equation*}
\verb!p <- P(eqnvec1, condition = "one") + P(eqnvec2, condition = "two")!. 
\end{equation*}
All symbols from \code{eqnvec1} and \code{eqnvec2} are collected and their union constitutes the symbols of $\theta$. The evaluation \code{p(theta)} returns a list of length $n$ (in the example $n = 2$) with inner parameters. The \code{"+"} operator can be applied consecutively to add condition by condition.

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)).
	\end{aligned}
	\label{}
\end{align*}
In words, if prediction- or observation functions are defined for different conditions then they expect condition-specific inputs which are evaluated by the matching functions. Examples for condition-specific prediction functions typically involve different forcings or events. Observation functions could be different if another  measurement technique has been used. In all these cases, the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined, referencing the condition, and combined by the \code{"+"} operator analogously to \code{p}.
% Aufruf der Funktionen mit subset von conditions

All commands, \code{P()}, \code{Xs()}, \code{Y()}, etc.~can be executed with \code{condition = NULL}. In that case, the corresponding returned function is generic and, if called for different conditions, the same identical function is evaluated with the condition-specific input. The other way round, if a function say \code{x} is defined for several conditions, the prediction can be evaluated for a subsect of conditions by \code{x(times, pars, conditions = myconditions)}.

\subsection{The data structure}

In \pkg{dMod} different experimental conditions are handled by lists. Parameter transformations, prediction- and observation functions stacked by the \code{"+"}

@@

The aim of \pkg{dMod} is parameter estimation. The objective function to be minimized for this purpose is the weighted least squares function which is produced by the command \code{normL2(data, prdfn)} where \code{data} is a \code{datalist} object and \code{prdfn} is a prediction function. A \code{datalist} is a list of \code{data.frame}s providing time points, observables, measured values and uncertainties for each condition. The \code{as.datalist()} command allows to construct a datalist from a \code{data.frame} as it is typically used for mixed effects models or plotting with \pkg{ggplot2}. % Satz verwirrend
